---
title: "ABDA Hiereichical"
author: "HamzaAfzalAshraf"
date: "2025-01-14"
output: html_document
---

```{r}
#Number of columns

colnames(read.csv("Student_performance_data.csv"))
```
```{r}
# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

```

```{r}
print(student_data$GradeClass)
sum(is.na(student_data$GradeClass)) 
```


```{r}
# Load required libraries
library(brms)


# Load the dataset (adjust the file path as necessary)
student_data <- read.csv("Student_performance_data.csv")



# check prior assumptions
table(student_data$Ethnicity)
table(student_data$ParentalEducation)

  

```
## inequal priors, means there's no exchangeability. Should work on fixing the priors.

```{r}
library(brms)
library(ggplot2)

```

```{r}
# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)
# Define the hierarchical model formula with nested effects
formula <- GradeClass ~ Age_z + StudyTimeWeekly_z + Absences_z + Gender + 
           Tutoring + ParentalSupport + Extracurricular + Sports + Music + Volunteering +
           (1 | Ethnicity/ParentalEducation)

# Specify priors
priors <- c(
  set_prior("normal(0, 2)", class = "Intercept"),       # Prior for intercept
  set_prior("normal(0, 1)", class = "b"),              # Prior for fixed effects
  set_prior("student_t(3, 0, 5)", class = "sd")        # Prior for random effects
)

# Fit the Bayesian hierarchical model
model <- brm(
  formula = formula,
  family = "cumulative",   # Ordered logistic regression for ordinal outcome
  data = student_data,
  prior = priors,
  chains = 4,              # Number of chains
  iter = 4000,             # Total iterations
  warmup = 2000,           # Warmup iterations
  control = list(adapt_delta = 0.99, max_treedepth = 15)  # Adjust control settings
)

# Print a summary of the model
summary(model)


# Visualize convergence and posterior distributions
plot(model)


```
# result analysis from the above: Students with higher study time and parental support are more likely to achieve better grades, while frequent absences and lack of extracurricular engagement are associated with poorer performance.
```{r}
pp_check(model)
```

```{r}
pairs(model, pars = c("b_Intercept", "sd_Intercept", "Age_z"))

```

#SAVE THE MODEL
```{r}
# Optional: Save the model for future use
saveRDS(model, "Bayesian_Student_Grade_First_Model.rds")
```

```{r}
# Convert GradeClass to a factor, maintaining the correct order: "A" (best) to "F" (worst)
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Convert GradeClass to numeric values (1 for A, 2 for B, etc.)
numeric_grades <- as.numeric(student_data$GradeClass)

# Ensure breaks span the range of both observed and posterior predictive data
breaks <- seq(0.5, 5.5, by = 1)  # Adjust to have bins between the grades

# Plot the histogram for observed grades
hist(numeric_grades, breaks = breaks, col = "black", 
     main = "Observed vs Posterior Predictive Grades",
     xlab = "Grade Class", ylab = "Frequency", xaxt = "n", freq = TRUE)

# Add custom labels for the x-axis corresponding to the correct grades
axis(1, at = 1:5, labels = c("F", "D", "C", "B", "A"))

# Overlay Posterior Predictive Samples (assuming you have 'y_rep')
for (i in 1:nrow(y_rep)) {
  # Assuming y_rep is already the correct numeric representation
  hist(y_rep[i, ], breaks = breaks, col = "red", 
       add = TRUE, freq = TRUE)
}

# Add a legend
legend("topright", legend = c("Observed Grades", "Posterior Predictive"), 
       fill = c("black", "red"))


# Print observed grades frequency
print("Observed Grades Frequency:")
observed_grades_table <- table(numeric_grades)
print(observed_grades_table)

# Print posterior predictive grades frequency
print("Posterior Predictive Grades Frequency:")
posterior_predictive_grades <- apply(y_rep, 2, function(x) floor(median(x)))
posterior_grades_table <- table(posterior_predictive_grades)
print(posterior_grades_table)

```



##Significant Predictors:

#StudyTimeWeekly_z (-0.62): Negative relationship with grade class, indicating that as study time increases, the probability of being in a higher grade class decreases (likely due to overcompensation by weaker students).
#Absences_z (3.33): Strong positive association, suggesting that higher absences significantly increase the likelihood of being in a lower grade class.
#Tutoring (-0.99): Negative effect, implying that students receiving tutoring are more likely to belong to a lower grade class (possibly because they already struggle academically).
#ParentalSupport (-0.59): Negative effect, indicating that less parental support is linked with lower grades.
#Extracurricular (-0.71): Participating in extracurricular activities correlates with a lower grade class, potentially reflecting time trade-offs.
#Sports (-0.75): Similarly, involvement in sports correlates with lower grades.
#Music (-0.45): Participation in music has a smaller but still negative association with grade class.


##Non-Significant Predictors:

#Age_z (0.02): Credible interval includes zero, suggesting no strong evidence for a relationship.
#Gender (-0.07): Minimal effect with CI including zero, indicating it may not be a meaningful predictor.
#Volunteering (0.14): Positive coefficient, but the CI includes zero, making it insignificant.


```{r}
# Histogram of StudyTimeWeekly
hist(student_data$StudyTimeWeekly, col = "purple", main = "Histogram of Study Time (Weekly)", 
     xlab = "Study Time (Hours)", ylab = "Frequency", breaks = 10)

```


```{r}
# Optional: Save the model for future use
saveRDS(model, "Bayesian_Student_Grade_Model_second.rds")

```





```{r}
# Extract the posterior predictive samples from the model
posterior_predictiv <- posterior_predict(model)



```




```{r}
#better grouping

# Load required libraries
library(brms)

# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)

# Create a new variable for activity group (Active or Inactive based on participation)
student_data$ActivityGroup <- with(student_data, 
                                    ifelse(Extracurricular == 1 | Sports == 1 | Music == 1 | Volunteering == 1, 
                                           "Active", "Inactive"))

# studytime on absences to know if studs are making up for lost time
formula <- GradeClass ~ StudyTimeWeekly_z * Absences_z + Gender + 
           ParentalSupport + Extracurricular + Sports + Music + Volunteering + 
           (1 + StudyTimeWeekly_z + Absences_z | Tutoring) + 
           (1 | ActivityGroup)  # Random intercept for ActivityGroup

# Specify priors
priors <- c(
  set_prior("normal(0, 2)", class = "Intercept"),       # Prior for intercept
  set_prior("normal(0, 1)", class = "b"),              # Prior for fixed effects
  set_prior("student_t(3, 0, 5)", class = "sd"),        # Prior for random effects
  set_prior("normal(0, 1)", class = "sd", group = "Tutoring"),  # Random effects for Tutoring
  set_prior("normal(0, 1)", class = "sd", group = "ActivityGroup")  # Random effects for ActivityGroup
)

# Fit the Bayesian hierarchical model
model <- brm(
  formula = formula,
  family = "cumulative",   # Ordered logistic regression for ordinal outcome
  data = student_data,
  prior = priors,
  chains = 4,              # Number of chains
  iter = 2000,             # Total iterations
  warmup = 1000,           # Warmup iterations
  control = list(adapt_delta = 0.99, max_treedepth = 15)  # Adjust control settings
)

# Print a summary of the model
summary(model)


```

```{r}
# Visualize convergence and posterior distributions
plot(model)


```

```{r}
# Optional: Save the model for future use
saveRDS(model, "Model2OrderedLogisticRegressionforOrdinal.rds")
```




```{r}
# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Convert GradeClass to GPA (continuous)
# Assuming you already have GPA in the dataset (if not, you need to calculate it)
# GPA is expected to be continuous (e.g., 0 to 4 scale)
# For this example, let's assume GPA is already present in the dataset

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)

# Create a new variable for activity group (Active or Inactive based on participation)
student_data$ActivityGroup <- with(student_data, 
                                    ifelse(Extracurricular == 1 | Sports == 1 | Music == 1 | Volunteering == 1, 
                                           "Active", "Inactive"))

# Multilevel formula for GPA as continuous
formula <- GPA ~ StudyTimeWeekly_z * Absences_z + Gender + 
           ParentalSupport + Extracurricular + Sports + Music + Volunteering + 
           (1 + StudyTimeWeekly_z + Absences_z | Tutoring) + 
           (1 | ActivityGroup)  # Random intercept for ActivityGroup

# Specify priors
priors <- c(
  set_prior("normal(0, 2)", class = "Intercept"),       # Prior for intercept
  set_prior("normal(0, 1)", class = "b"),               # Prior for fixed effects
  set_prior("student_t(3, 0, 5)", class = "sd"),        # Prior for random effects
  set_prior("normal(0, 1)", class = "sd", group = "Tutoring"),  # Random effects for Tutoring
  set_prior("normal(0, 1)", class = "sd", group = "ActivityGroup")  # Random effects for ActivityGroup
)

# Fit the Bayesian hierarchical model using Gaussian family for continuous GPA
model_continuous <- brm(
  formula = formula,
  family = "gaussian",   # Gaussian family for continuous outcome
  data = student_data,
  prior = priors,
  chains = 4,              # Number of chains
  iter = 2000,             # Total iterations
  warmup = 1000,           # Warmup iterations
  control = list(adapt_delta = 0.99, max_treedepth = 15)  # Adjust control settings
)

# Print a summary of the model
summary(model_continuous)

```
```{r}
# Optional: Save the model for future use
saveRDS(model_continuous, "Model1GaussianForContinuous.rds")
```

```{r}
summary(model_continuous)

```


```{r}
# Generate posterior predictive samples
pp_samples <- posterior_predict(model_continuous)

# Visualize PPC
library(bayesplot)
pp_check(model_continuous, ndraws = 100) # Overlay simulated data with observed data

# Histograms of predictions
pp_check(model_continuous, type = "hist", ndraws = 100)

# Compare means or other summary statistics
pp_check(model_continuous, type = "stat", stat = "mean")
pp_check(model_continuous, type = "stat", stat = "sd")

```



```{r}
# Generate posterior predictive samples
pp_samples <- posterior_predict(model)

# Visualize PPC
library(bayesplot)
pp_check(model, ndraws = 100) # Overlay simulated data with observed data

# Histograms of predictions
pp_check(model, type = "hist", ndraws = 100)

# Compare means or other summary statistics
pp_check(model, type = "stat", stat = "mean")
pp_check(model, type = "stat", stat = "sd")

```

```{r}
library(brms)
# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Preprocessing Steps

# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)

# Create a new variable for activity group (Active or Inactive)
student_data$ActivityGroup <- with(student_data, 
                                    ifelse(Extracurricular == 1 | Sports == 1 | Music == 1 | Volunteering == 1, 
                                           "Active", "Inactive"))

# Ensure ActivityGroup is a factor
student_data$ActivityGroup <- factor(student_data$ActivityGroup)

# Model Specification

# Define refined priors (without the disc parameter)
refined_priors <- c(
  prior(normal(0.2, 0.1), class = "b", coef = "StudyTimeWeekly_z"),
  prior(normal(-1.0, 0.5), class = "b", coef = "Absences_z"),
  prior(normal(0, 0.5), class = "b", coef = "StudyTimeWeekly_z:Absences_z"),
  prior(normal(0, 0.1), class = "b", coef = "Gender"),
  prior(normal(0.5, 0.1), class = "b", coef = "ParentalSupport"),
  prior(normal(0.5, 0.2), class = "b", coef = "Extracurricular"),
  prior(normal(0.5, 0.2), class = "b", coef = "Sports"),
  prior(normal(0.3, 0.2), class = "b", coef = "Music"),
  prior(normal(0, 0.2), class = "b", coef = "Volunteering"),
  prior(student_t(3, 0, 1), class = "sd", group = "ActivityGroup"),
  prior(student_t(3, 0, 2), class = "sd", group = "Tutoring"),
  prior(lkj(2), class = "cor", group = "Tutoring"),
  prior(normal(0, 2), class = "Intercept")
)

# Fit the model
refined_model <- brm(
  formula = GradeClass ~ StudyTimeWeekly_z * Absences_z + Gender + ParentalSupport + 
             Extracurricular + Sports + Music + Volunteering + 
             (1 + StudyTimeWeekly_z + Absences_z | Tutoring) + 
             (1 | ActivityGroup),
  data = student_data,
  family = cumulative(link = "logit"),
  prior = refined_priors,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  cores = 4
)

# Model Summary
summary(refined_model)

```

```{r}
# Run PPC on the refined model
library(bayesplot)

# Overlay plot of observed vs simulated data
pp_check(refined_model, ndraws = 100)

# Histograms of predicted vs observed
pp_check(refined_model, type = "hist", ndraws = 100)

# Compare means or other summary statistics
pp_check(refined_model, type = "stat", stat = "mean")
pp_check(refined_model, type = "stat", stat = "sd")


```




```{r}
# Optional: Save the model for future use
saveRDS(refined_model, "RedefinedModel12OrderedLogisticRegressionforOrdinal.rds")
```



```{r}
# Load required libraries
library(brms)
library(bayesplot)

# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Ensure that the data is correctly processed
# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)

# Create a new variable for activity group (Active or Inactive based on participation)
student_data$ActivityGroup <- with(student_data, 
                                    ifelse(Extracurricular == 1 | Sports == 1 | Music == 1 | Volunteering == 1, 
                                           "Active", "Inactive"))

# Define priors
refined_priors <- c(
  set_prior("normal(0, 1)", class = "b"),         # Fixed effects
  set_prior("normal(0, 5)", class = "Intercept"), # Intercept
  set_prior("student_t(3, 0, 2.5)", class = "sd"),# Random effects standard deviation
  set_prior("lkj(2)", class = "cor")             # Correlation matrices
)

# Fit the refined model
refined_model <- brm(
  formula = GradeClass ~ StudyTimeWeekly_z * Absences_z + Gender + ParentalSupport + 
             Extracurricular + Sports + Music + Volunteering + 
             (1 + StudyTimeWeekly_z + Absences_z | Tutoring) + 
             (1 | ActivityGroup),
  data = student_data,
  family = cumulative(link = "logit"),
  prior = refined_priors,
  iter = 2000,
  warmup = 1000,
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.95)
)

# Check model summary
summary(refined_model)

```

```{r}
# Check model summary
summary(refined_model)

# Perform posterior predictive checks
pp_check(refined_model, ndraws = 100)

# Generate posterior predictions
posterior_samples <- posterior_predict(refined_model, ndraws = 100)
head(posterior_samples)

# Conduct posterior predictive checks (PPC)
pp_check(refined_model)

# Check for divergent transitions
rstan::check_divergences(refined_model$fit)

# Visualize the random effects
plot(refined_model, pars = "^sd_")

# Check prior summary
prior_summary(refined_model)


```


```{r}

# Load required libraries
library(brms)
library(bayesplot)

# Read in your dataset
student_data <- read.csv("Student_performance_data.csv")

# Ensure that the data is correctly processed
# Convert GradeClass to an ordered factor
student_data$GradeClass <- factor(student_data$GradeClass, 
                                  levels = c(0, 1, 2, 3, 4), 
                                  labels = c("A", "B", "C", "D", "F"), 
                                  ordered = TRUE)

# Scale continuous predictors
student_data$Age_z <- scale(student_data$Age)
student_data$StudyTimeWeekly_z <- scale(student_data$StudyTimeWeekly)
student_data$Absences_z <- scale(student_data$Absences)

# Create a new variable for activity group (Active or Inactive based on participation)
student_data$ActivityGroup <- with(student_data, 
                                    ifelse(Extracurricular == 1 | Sports == 1 | Music == 1 | Volunteering == 1, 
                                           "Active", "Inactive"))


refined_priors <- c(
  set_prior("normal(0, 1)", class = "b"),         # Fixed effects
  set_prior("normal(0, 5)", class = "Intercept"), # Intercept
  set_prior("normal(0, 1)", class = "sd"),        # Random effects standard deviation
  set_prior("lkj(2)", class = "cor")             # Correlation matrices
)

refined_model <- brm(
  formula = GradeClass ~ StudyTimeWeekly_z + Absences_z + Gender + ParentalSupport + 
             Extracurricular + Sports + Music + Volunteering + 
             (1 | Tutoring) + 
             (1 | ActivityGroup),
  data = student_data,
  family = cumulative(link = "logit"),
  prior = refined_priors,
  iter = 3000,
  warmup = 1000,
  chains = 4,
  cores = 4,
  control = list(adapt_delta = 0.99, max_treedepth = 15)
)


```


